The large scale clustering of fast radio bursts in dispersion measure space

### 2015-03-19 writeup/dm_space_distortions.tex
Many changes.

\section{FRB density perturnbations in DM space}

\begin{equation}
D(\hat n, \chi) = \int_0^\chi \ud \chi' a(\chi')^2
   \left[1 + \delta_e(\hat{n}\chi')\right]
\end{equation}

This defines DM space:
\begin{equation}
D(\chi_s) =  \int_0^{\chi_s} \ud \chi' a(\chi')^2
\end{equation}

@@

The jacobian in spherical coordinates is:
\begin{align}
\left| \frac{\ud^3\vec x}{\ud^3\vec x_s} \right|
    &= \frac{\ud \chi}{\ud \chi_s}\frac{\chi^2}{\chi_s^2}\\
    &= (1 + \delta_e)^{-1}
       \left(1 + \frac{\int_0^\chi \ud \chi' \delta_e(\hat n \chi')}
                      {\chi}\right)^{-2}\\
    &\approx 1 - \delta_e
        - \frac{2}{\chi}\int_0^\chi \ud \chi' \delta_e(\hat n \chi')
        \label{e:jac}
\end{align}
Following Kaiser, the third line follows since the
second factor in parenthases on the second line, which is the
average overdensity along the line of sight, is much
closer to unity than the first due to cancelations.

@@

In this section all quantities are already first order in the perturbations so
there is no need to distinguish between the DM space coordinates and the real
space coordintes. The correction is higher order. Still need to distinguish
between DM space and real space fields ($\delta_{fs} \neq \delta_f$, but
$\delta_f(\vec x_s) \approx \delta_f(\vec x)$).

@@

Because the expression for the density perturbations in DM space involves an
integral over the line of sight, which for far off sources may cover a
significant amount of the Universe's expansion, it is nessisary to explicitly
account for light cone effects. This is most easily done in linear theory,
where these effects are scale independant, a factorization that allows certian
integral to be performed. In reality, non-linear growth causes both the bias
and the growth factor to be scale and time dependant. As in weak lensing, this
would need to be simulated.

We use the following convensions:
\begin{equation}
    P_X(k, \chi) = b_X^2 g(\chi)^2 a(\chi)^2 P(k, 0)
\end{equation}

\begin{equation}
    \delta_X(\vec x, \chi) =  b_X g(\chi) a(\chi) \int\frac{\ud^3\vec k}{(2 \pi)^3}
        e^{i\vec k \cdot \vec x} \delta(\vec k, 0)
\end{equation}

\begin{equation}
    w^{XX}(\hat n \cdot \hat n', \chi, \chi')
    \equiv \langle \delta_X(\vec x) \delta_X(\vec x') \rangle
\end{equation}

For dark matter:
\begin{align}
w(\hat n \cdot \hat n', \chi, \chi') 
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} \frac{\ud^3\vec k'}{(2 \pi)^3}
        e^{i\vec k \cdot \vec x} e^{i\vec k' \cdot \vec x'}
        \langle \delta(\vec k, 0) \delta(\vec k', 0) \rangle\\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i\vec k \cdot (\vec x - \vec x')} P(k, 0)\\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i\vec k \cdot (\vec x - \vec x')} P(k)\\
    &= gg'aa'
        \int\frac{k^2 \sin(\theta_k) \ud k \ud \theta_k \ud \phi_k}{(2 \pi)^3} 
        e^{i k |\vec x - \vec x'| \cos\theta_k} P(k, 0)\\
    &= gg'aa'
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        \frac{\sin(k |\vec x - \vec x'|)}{k |\vec x - \vec x'|} P(k, 0)\\
    &= gg'aa'
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        j_0 (k \sqrt{\chi^2 + \chi'^{^2} - \chi \chi' \hat n \cdot \hat n'})
        P(k, 0)
\end{align}
Somehow when I integrate this against a lengendre polynomial, $P_\ell(\mu)$,
to get $C_\ell$
I'm going to get Bessel functions. Lets assume that works (Kris, how does this
work?) to get:
\begin{equation}
    C_{\ell\chi\chi'} = \frac{2 gg'aa'}{\pi}
    \int_0^\infty\ud k k^2 j_\ell(k \chi) j_\ell(k \chi')
        P(k, 0)
\end{equation}

Define $\delta_d$ as:
\begin{equation}
    \delta_d(\hat n \chi) = \int_0^\chi \ud \chi' \delta_e(\hat n \chi')
\end{equation}
which doesn't have the same units as overdensity, but that is okay.
\begin{align}
    w^{dd}(\hat n \cdot \hat n', \chi, \chi')
    &= \int_0^\chi \ud \chi'' \int_0^{\chi'} \ud \chi'''
         \langle \delta_e(\hat n \chi'') \delta_e(\hat n' \chi''') \rangle\\
    &= b_e^2\int_0^\chi \ud \chi'' \int_0^{\chi'} \ud \chi'''
        g''g'''a''a'''
        \int\frac{\ud^3\vec k}{(2 \pi)^3} \frac{\ud^3\vec k'}{(2 \pi)^3}
        e^{i\vec k \cdot \vec x''} e^{i\vec k' \cdot \vec x'''}
        \langle \delta(\vec k, 0) \delta(\vec k', 0) \rangle\\
    &= b_e^2\int_0^\chi \ud \chi'' \int_0^{\chi'} \ud \chi'''
        g''g'''a''a'''
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        j_0 (k \sqrt{\chi''^2 + \chi'''^{^2} - \chi'' \chi''' \hat n \cdot \hat n'})
        P(k, 0)
\end{align}

>>>>

\section{FRB density perturbations in DM space}

\begin{equation}
D(\hat n, \chi) = \int_0^\chi \frac{\ud\chi'}{a(\chi')}
   \left[1 + \delta_e(\hat{n}\chi')\right]
\end{equation}

This defines DM space:
\begin{equation}
D(\chi_s) = \int_0^{\chi_s}\frac{\ud\chi'}{a(\chi')}
\end{equation}

@@

The jacobian in spherical coordinates is:
\begin{align}
\left| \frac{\ud^3\vec x}{\ud^3\vec x_s} \right|
    &= \frac{\ud \chi}{\ud \chi_s}\frac{\chi^2}{\chi_s^2}\\
    &= (1 + \delta_e)^{-1}
       \left(1 + \frac{\int_0^\chi \ud \chi' \delta_e(\hat n \chi')}
                      {\chi}\right)^{-2}\\
    &\approx 1 - \delta_e
        - \frac{2}{\chi}\int_0^\chi \ud \chi' \delta_e(\hat n \chi')
        \label{e:jac}
\end{align}
Following Kaiser, the third line follows since the
second factor in parentheses on the second line, which is the
average overdensity along the line of sight, is much
closer to unity than the first due to cancellations.

@@

In this section all quantities are already first order in the perturbations so
there is no need to distinguish between the DM space coordinates and the real
space coordinates. The correction is higher order. Still need to distinguish
between DM space and real space fields ($\delta_{fs} \neq \delta_f$, but
$\delta_f(\vec x_s) \approx \delta_f(\vec x)$).

@@

Because the expression for the density perturbations in DM space involves an
integral over the line of sight, which for far off sources may cover a
significant amount of the Universe's expansion, it is necessary to explicitly
account for light cone effects. This is most easily done in linear theory,
where these effects are scale independent, a factorization that allows certain
integral to be performed. In reality, non-linear growth causes both the bias
and the growth factor to be scale and time dependant. As in weak lensing, this
would need to be simulated.

We use the following conventions:
\begin{equation}
    P_{xy}(k, \chi) = b_x b_y g(\chi)^2 a(\chi)^2 P(k, 0)
\end{equation}

\begin{equation}
    \delta_x(\vec x, \chi) =  b_x g(\chi) a(\chi) \int\frac{\ud^3\vec k}{(2 \pi)^3}
        e^{i\vec k \cdot \vec x} \delta(\vec k, 0)
\end{equation}

\begin{equation}
    w^{xy}(\hat n \cdot \hat n', \chi, \chi')
    \equiv \langle \delta_x(\vec x) \delta_y(\vec x') \rangle
\end{equation}

For dark matter, ignoring time dependance (redshift evolution)
of the power spectrum:
\begin{align}
\delta_{\ell\ell'}\delta_{mm'}C_\ell(\chi, \chi')
    &= \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
        \langle \delta(\hat n \chi) \delta(\hat n' \chi') \rangle
        \\
    &= gg'aa'
        \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
        \int\frac{\ud^3\vec k}{(2 \pi)^3} \frac{\ud^3\vec k'}{(2 \pi)^3}
        e^{i\vec k \cdot \vec x} e^{i\vec k' \cdot \vec x'}
        \langle \delta(\vec k, 0) \delta(\vec k', 0) \rangle
        \\
    &= gg'aa'
        \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i\vec k \cdot (\vec x - \vec x')} P(k, 0)
        \\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} P(k, 0)
        \int\ud\Omega Y_{\ell m}(\hat n)
        e^{i\vec k \cdot \vec x}
        \int\ud\Omega' Y_{\ell' m'}(\hat n')
        e^{-i\vec k \cdot \vec x'}
        \\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} P(k, 0)
        \left[(4 \pi) i^\ell j_\ell(k\chi) Y_{\ell m}^*(\hat k)\right]
        \left[(4 \pi) i^{-\ell'} j_{\ell'}(k\chi') Y_{\ell' m'}(\hat
        k)\right]
        \\
    &= gg'aa'\frac{2}{\pi}
        \int\ud k k^2 j_\ell(k\chi) j_{\ell'}(k\chi')P(k, 0)
        \int\ud \Omega_k i^{(\ell - \ell')}
        Y_{\ell' m'}^*(\hat k) Y_{\ell m}(\hat k)
        \\
    &= \delta_{\ell\ell'}\delta_{mm'}gg'aa'\frac{2}{\pi}
        \int\ud k k^2  j_\ell(k\chi) j_{\ell}(k\chi')P(k, 0)
        \\
C_\ell(\chi,\chi') 
    &= gg'aa'\frac{2}{\pi}
        \int\ud k k^2 j_\ell(k\chi) j_{\ell}(k\chi')P(k, 0)
\end{align}

\section{Small angles and Limber}

Alternately, on the flat sky and for small angles:
\begin{align}
(2\pi)^2\delta^2(\vec \ell - \vec \ell')C_\ell(\chi, \chi')
    &=
        \int\ud^2\vec\theta\ud^2\vec\theta'
        e^{-i\vec\ell\cdot\vec\theta} e^{-i\vec\ell'\cdot\vec\theta'}
        \langle \delta(\hat n \chi) \delta(\hat n' \chi') \rangle
        \\
    &= gg'aa'
        \int\ud^2\vec\theta\ud^2\vec\theta'
        e^{-i\vec\ell\cdot\vec\theta} e^{-i\vec\ell'\cdot\vec\theta'}
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i\vec k \cdot (\vec x - \vec x')} P(k, 0)
        \\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i k_\parallel (\chi - \chi')} P(k, 0)
        \int\ud^2\vec\theta e^{-i(\vec\ell-\bar\chi\vec k_\bot)\cdot\vec\theta}
        \int\ud^2\vec\theta' e^{-i(\vec\ell'-\bar\chi\vec k_\bot)\cdot\vec\theta'}
        \\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i k_\parallel (\chi - \chi')} P(\sqrt{k_\parallel^2 + k_\bot^2}, 0)
        (2\pi)^4 \delta^2(\vec\ell - \bar\chi \vec k_\bot)
        \delta^2(\vec\ell' - \bar\chi\vec k_\bot)
        \\
    &= gg'aa'
        \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
        e^{i k_\parallel (\chi - \chi')}
        P(\sqrt{k_\parallel^2 + \ell^2/\bar\chi^2}, 0)
        \frac{(2\pi)^2\delta^2(\vec\ell' - \vec \ell)}{\bar\chi^2}
        \\
C_\ell(\chi, \chi')
    &= \frac{gg'aa'}{\bar\chi^2}
        \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
        e^{i k_\parallel (\chi - \chi')}
        P(\sqrt{k_\parallel^2 + \ell^2/\bar\chi^2}, 0)
\end{align}
Which is valid for $\ell \gg 1$ and $\chi - \chi' \ll \bar\chi$.
$\bar \chi \equiv (\chi + \chi') /2$

Define $\delta_d$ as:
\begin{equation}
    \delta_d(\hat n \chi) = \int_0^\chi \ud \chi' \delta_e(\hat n \chi')
\end{equation}
which doesn't have the same units as over-density, but that is okay.
\begin{equation}
C^{dd}_\ell(\chi,\chi') 
    =
    \frac{2}{\pi}
    \int_0^\chi\ud\chi''g''a''
    \int_0^{\chi'}\ud\chi'''
    g'''a'''
    \int\ud k k^2 j_\ell(k\chi'') j_{\ell}(k\chi''')P_{ee}(k, 0)
\end{equation}
At this point we could again to the transformation to $C_{\ell\chi\chi'}$
leaving the expression as a triple integral. This is not as computationally
costly as might be expected at first glance, since the $\chi''$ and $\chi'''$
integrals are almost the same for all $\chi$, $\chi'$ pairs with only the
limits of integration changing. Thus the full set of $C_{\ell\chi\chi'}$'s can
be computed with a cumulative sum, only performing one new integral (over $k$)
for each element.

Or for flat sky and small angles:
\begin{align}
C^{dd}_\ell(\chi,\chi') 
    &=
    \int_0^\chi\ud\chi''g''a''
    \int_0^{\chi'}\ud\chi'''g'''a'''
    \frac{1}{\bar\chi^2}
    \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
    e^{i k_\parallel (\chi'' - \chi''')}
    P_{ee}(\sqrt{k_\parallel^2 + \ell^2/\bar\chi^2}, 0)
    \\
    &\approx
    \int_0^\chi\ud\chi''g''a''
    \int_0^{\chi'}\ud\chi'''g'''a'''
    \frac{1}{\bar\chi^2}
    \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
    e^{i k_\parallel (\chi'' - \chi''')}
    P_{ee}(\ell/\bar\chi, 0)
    \\
    &=
    \int_0^\chi\ud\chi''g''a''
    \int_0^{\chi'}\ud\chi'''g'''a'''
    \frac{1}{\bar\chi^2}
    \delta(\chi'' - \chi''')
    P_{ee}(\ell/\bar\chi, 0)
    \\
    &=
    \int_0^{\min(\chi,\chi')}\ud\chi''
    \left(\frac{g''a''}{\chi''}\right)^2
    P_{ee}(\ell/\chi'', 0)
\end{align}
\red{Double check the $\min$.}
where the second line applies Limber's approximation.

Finally, the cross terms:
\begin{equation}
C^{dx}_\ell(\chi,\chi')
    =
    g'a'
    \frac{2}{\pi}
    \int_0^\chi\ud\chi''g''a''
    \int\ud k k^2 j_\ell(k\chi'') j_{\ell}(k\chi')P_{ex}(k, 0)
\end{equation}
Or under Limber:
\begin{align}
C^{dx}_\ell(\chi,\chi')
    &=
    g'a'
    \int_0^\chi\ud\chi''g''a''
    \frac{1}{\bar\chi^2}
    \delta(\chi'' - \chi')
    P_{ex}(\ell/\bar\chi, 0)
    \\
    &=
    \left(\frac{g'a'}{\chi'}\right)^2
    P_{ex}(\ell/\chi', 0)
    \qquad \textrm{if $\chi' < \chi$ and 0 otherwise.}
    \\
C^{dx}_\ell(\chi,\chi') + C^{xd}_\ell(\chi,\chi')
    &=
    \left(\frac{g^ma^m}{\chi^m}\right)^2
    P_{ex}(\ell/\chi^m, 0)
\end{align}

Accumulating all terms for Limber and assuming a linear bias model:
\begin{align}
C^{ss}_\ell(\chi_m,\Delta\chi) =&~
    (b_f - b_e)^2 \left(\frac{g_m a_m}{\chi_m}\right)^2
        \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
        e^{i k_\parallel \Delta\chi}
        P(\sqrt{k_\parallel^2 + \ell^2/\chi_m^2}, 0)
    \\ &+
    \left(\frac{1}{\bar{n}_f}\frac{\ud \bar{n}_f}{\ud \chi}
    + \frac{2}{\chi} \right)_{\chi_m}^2
    b_e^2 \int_0^{\chi_m}\ud\chi''
        \left(\frac{g''a''}{\chi''}\right)^2
        P(\ell/\chi'', 0)
    \\ &+
    \left(\frac{1}{\bar{n}_f}\frac{\ud \bar{n}_f}{\ud \chi}
    + \frac{2}{\chi} \right)_{\chi_m}
    b_e(b_f - b_e)\left(\frac{g_m a_m}{\chi_m}\right)^2
        P(\ell/\chi_m, 0)
\end{align}

%%%% SCRAP

\begin{comment}
\begin{align}
w(\hat n \cdot \hat n', \chi, \chi') 
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} \frac{\ud^3\vec k'}{(2 \pi)^3}
        e^{i\vec k \cdot \vec x} e^{i\vec k' \cdot \vec x'}
        \langle \delta(\vec k, 0) \delta(\vec k', 0) \rangle\\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i\vec k \cdot (\vec x - \vec x')} P(k, 0)\\
    &= gg'aa'
        \int\frac{\ud^3\vec k}{(2 \pi)^3} 
        e^{i\vec k \cdot (\vec x - \vec x')} P(k)\\
    &= gg'aa'
        \int\frac{k^2 \sin(\theta_k) \ud k \ud \theta_k \ud \phi_k}{(2 \pi)^3} 
        e^{i k |\vec x - \vec x'| \cos\theta_k} P(k, 0)\\
    &= gg'aa'
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        \frac{\sin(k |\vec x - \vec x'|)}{k |\vec x - \vec x'|} P(k, 0)\\
    &= gg'aa'
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        j_0 (k \sqrt{\chi^2 + \chi'^{^2} - \chi \chi' \hat n \cdot \hat n'})
        P(k, 0)
\end{align}
Somehow when I integrate this against a Legendre polynomial, $P_\ell(\mu)$,
to get $C_\ell$
I'm going to get Bessel functions. Lets assume that works (Kris, how does this
work?) to get:
\begin{equation}
    C_{\ell\chi\chi'} = \frac{2 gg'aa'}{\pi}
    \int_0^\infty\ud k k^2 j_\ell(k \chi) j_\ell(k \chi')
        P(k, 0)
\end{equation}
\end{comment}

\begin{comment}
\begin{align}
    w^{dd}(\hat n \cdot \hat n', \chi, \chi')
    &= \int_0^\chi \ud \chi'' \int_0^{\chi'} \ud \chi'''
         \langle \delta_e(\hat n \chi'') \delta_e(\hat n' \chi''') \rangle\\
    &= b_e^2\int_0^\chi \ud \chi'' \int_0^{\chi'} \ud \chi'''
        g''g'''a''a'''
        \int\frac{\ud^3\vec k}{(2 \pi)^3} \frac{\ud^3\vec k'}{(2 \pi)^3}
        e^{i\vec k \cdot \vec x''} e^{i\vec k' \cdot \vec x'''}
        \langle \delta(\vec k, 0) \delta(\vec k', 0) \rangle\\
    &= b_e^2\int_0^\chi \ud \chi'' \int_0^{\chi'} \ud \chi'''
        g''g'''a''a'''
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        j_0 (k \sqrt{\chi''^2 + \chi'''^{^2} - \chi'' \chi''' \hat n \cdot \hat n'})
        P(k, 0)
\end{align}

Alternately, we can make the limber approximation (replacing
$a(\chi'')g(\chi'')$ and $a(\chi''')g(\chi''')$ with 
$a(\bar \chi)g(\bar \chi)$) and small angle approximations
(replacing the argument of the spherical Bessel with
$k\sqrt{\Delta \chi^2 + (\bar \chi (\hat n - \hat n'))^2}$).
\begin{align}
    w^{dd}(\hat n \cdot \hat n', \chi, \chi')
    &= b_e^2\int_0^\chi \ud \chi'' \int_0^{\chi'} \ud \chi'''
        (g(\bar \chi) a(\bar \chi))^2
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        j_0 (k \sqrt{\Delta \chi^2 + (\bar \chi (\hat n - \hat n'))^2})
        P(k, 0)\\
    &= b_e^2\int_0^{(\chi + \chi')/2} \ud \bar\chi 
        (g(\bar \chi) a(\bar \chi))^2
        \int_0^\infty\frac{\ud k}{2 \pi^2} k^2
        P(k, 0)
        \int_{-\infty}^{\infty} \ud \Delta\chi
        j_0 (k \sqrt{\Delta \chi^2 + (\bar \chi (\hat n - \hat n'))^2})\\
\end{align}

\end{comment}
